---
title: "Recommendations for Using summarytools With Rmarkdown"
author: "Dominic Comtois"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recommendations for Rmarkdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

<!-- (output: html_document or html_vignette) -->

```{r setup, include=FALSE}
library(summarytools)
library(knitr)
opts_chunk$set(comment=NA, prompt=FALSE, cache=FALSE, echo=TRUE)
```

This document uses theme `rmarkdown::html_vignette`.

Below are examples using recommended styles for `rmarkdown` rendering. Available styles are:  

- _simple_ (default)  
- _rmarkdown_  
- _grid_  
- _multiline_  
 
Jump to...

- [descr()](#descr)  
- [ctable()](#ctable)  
- [dfSummary()](#dfsummary)  


--------------------------------------------------------------------------------

# freq()

`freq()` is best used with _default_ or _rmarkdown_ styles. HTML rendering is not recommended.

### Default (_simple_) Style
```{r freq_default, results='asis'}
freq(tobacco$gender)
```

### Default (_simple_) Style with plain.ascii = FALSE
```{r freq_default2, results='asis'}
freq(tobacco$gender, plain.ascii = FALSE)
```

## Rmarkdown Style
```{r freq_rm, results='asis'}
freq(tobacco$gender, style = 'rmarkdown')
freq(tobacco$gender, style = 'rmarkdown', omit.headings = TRUE)
```

<!-- ### Grid Style -->
<!-- ```{r freq_grid, results='asis'} -->
<!-- freq(tobacco$gender, style = 'grid') -->
<!-- ``` -->

<!-- ## HTML Rendering -->
<!-- ```{r freq_html, results='asis'} -->
<!-- print(freq(tobacco$gender), method = 'render') -->
<!-- print(freq(tobacco$gender), method = 'render', omit.headings = TRUE) -->
<!-- ``` -->

 
------------------------------------------------------------------------------------
 
<a href="#top">Back to top</a>
 
# descr() {#descr}
`descr()` is best used with _default_ or _rmarkdown_ styles. HTML rendering is also supported.

### Default Style
```{r descr_default, results='asis'}
descr(tobacco)
```

### Default Style with plain.ascii = FALSE
```{r descr_default2, results='asis'}
descr(tobacco, plain.ascii = FALSE)
descr(tobacco$BMI, plain.ascii = FALSE)
```

## Rmarkdown Style
```{r descr_rm, results='asis'}
descr(tobacco, style = 'rmarkdown')
descr(tobacco$BMI, style = 'rmarkdown')
descr(tobacco$BMI, style = 'rmarkdown', omit.headings = TRUE)
```

<!-- ### Grid Style -->
<!-- ```{r descr_grid, results='asis'} -->
<!-- descr(tobacco, style = 'grid') -->
<!-- descr(tobacco$BMI, style = 'grid') -->
<!-- ``` -->

## HTML Render
```{r descr_html, results='asis'}
print(descr(tobacco), method = 'render')
print(descr(tobacco$BMI), method = 'render')
print(descr(tobacco, omit.headings = TRUE), method = 'render')
```

 
--------------------------------------------------------------------------------

<a href="#top">Back to top</a>

# Ctable() {#ctable}

For best results, use *HTML rendering*.
<!-- ## Default Style -->
<!-- ```{r ctable_default, results='asis'} -->
<!-- ctable(tobacco$gender, tobacco$smoker) -->
<!-- ``` -->

<!-- ## Default Style with plain.ascii = FALSE -->
<!-- ```{r ctable_default2, results='asis'} -->
<!-- ctable(tobacco$gender, tobacco$smoker, plain.ascii = FALSE) -->
<!-- ``` -->

<!-- ## Rmarkdown Style -->
<!-- ```{r ctable_rm, results='asis'} -->
<!-- ctable(tobacco$gender, tobacco$smoker, style = 'rmarkdown') -->
<!-- ``` -->

<!-- ## Grid Style -->
<!-- ```{r ctable_grid, results='asis'} -->
<!-- ctable(tobacco$gender, tobacco$smoker, justify = 'right', style = 'grid') # voir si default est bien right -->
<!-- ``` -->

```{r ctable_html, results='asis'}
print(ctable(tobacco$gender, tobacco$smoker), method = 'render')
print(ctable(tobacco$gender, tobacco$smoker, omit.headings = TRUE), method = 'render')
```

--------------------------------------------------------------------------------

<a href="#top">Back to top</a>

# dfSummary() {#dfsummary}

For best results, use _grid_ style and set `graph.col = FALSE`. 

<!-- ## Default Style -->
<!-- ```{r dfs_default, results='asis'} -->
<!-- dfSummary(tobacco) -->
<!-- dfSummary(tobacco, graph.col = FALSE) -->
<!-- ``` -->

<!-- ## Default Style with plain.ascii = FALSE -->
<!-- ```{r dfs_default2, results='asis'} -->
<!-- dfSummary(tobacco, plain.ascii = FALSE) -->
<!-- dfSummary(tobacco, graph.col = FALSE, plain.ascii = FALSE) -->
<!-- ``` -->

<!-- ## Multiline Style -->
<!-- ```{r dfs_ml, results='asis'} -->
<!-- dfSummary(tobacco, style = 'multiline') -->
<!-- dfSummary(tobacco, style = 'multiline', graph.col = FALSE) -->
<!-- dfSummary(tobacco, style = 'multiline', plain.ascii = FALSE) -->
<!-- dfSummary(tobacco, style = 'multiline', plain.ascii = FALSE, graph.col = FALSE) -->
<!-- ``` -->

<!-- ## Grid Style -->
<!-- ```{r dfs_grid, results='asis'} -->
<!-- dfSummary(tobacco, style = 'grid') -->
<!-- dfSummary(tobacco, style = 'grid', graph.col = FALSE) -->
<!-- dfSummary(tobacco, style = 'grid', plain.ascii = FALSE) -->
<!-- ``` -->

## Grid Style
```{r dfs_grid, results='asis'}
dfSummary(tobacco, style = 'grid', plain.ascii = FALSE, graph.col = FALSE)
dfSummary(tobacco, style = 'grid', plain.ascii = FALSE, graph.col = FALSE, omit.headings = TRUE)
```

<!-- ## HTML Render -->
<!-- ```{r dfs_html, results='asis'} -->
<!-- print(dfSummary(tobacco), method = 'render') -->
<!-- ``` -->

<a href="#top">Back to top</a>

