---
title: "summarytools in the Age of Tidyverse"
author: "Dominic Comtois"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette: 
    css: 
    - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
vignette: >
  %\VignetteIndexEntry{Introduction to summarytools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{magrittr}
  %\VignetteDepends{forcats}
  %\VignetteDepends{dplyr}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(comment = NA, prompt = FALSE, cache = FALSE, results = 'asis')
library(summarytools)
library(dplyr)
library(forcats)
library(magrittr)
st_options(plain.ascii = FALSE,
           style = "rmarkdown",
           footnote = NA,
           subtitle.emphasis = FALSE)
```

```{r, echo=FALSE}
st_css()
```

# Introduction

I started writing **summarytools** several years ago (the first CRAN 
release dates back to 2014), way before the
[**tidyverse**](https://cran.r-project.org/package=tidyverse) came about,
profoundly impacting the way people use R, write their code and
organize their thoughts while doing work on and with their data. My
focus in the last months has in been making **summarytools** fit more
naturally in this framework, first by supporting pipe operators first from
[**magrittr**](https://cran.r-project.org/package=magrittr), and more
recently from [**pipeR**](https://cran.r-project.org/package=pipeR).

### Several Possible Avenues

I've considered making "twin copies" of the four core functions
so that one set would work best with the traditionnal / _base R_, and the
other with the **tidyverse**. I'Ve opted against that route
for various reasons, but mainly because I found it worth the challenge to
try and keep things unified, adding flexibility all the while maintaining 
full backward compatibility. This document is not an introduction to 
**summarytools** per se -- there is already an 
[introductory vignette](https://cran.r-project.org/package=summarytools/vignettes/Introduction.html) 
and [another vignette](https://cran.r-project.org/package=summarytools/vignettes/Recommendations-rmarkdown.html) 
focusing on the integration of **summarytools** outputs in _Rmarkdown_ 
documents. Rather, it is a more general exposé of the package's new 
functionalities seen through the lens of this dichotomy that has taken place
inside the _R_ language itself.

Still, a review of some of the basic basic functionalities is included, so
that readers with no prior knowledge of **summarytools** can also follow.

## The Four Core Function

The following functions constitute the foundations of the package. Having worked
in fields where the two software suites _SAS_ and _SPSS_ were largely 
predominant, I found _R_ frustrating at first, because I was so used to 
producing visually appealing univariate or bi-variate statistical outputs of
the following nature:

|                Category             |  _summarytools_ function |  _SAS_ procedure         |  _SPSS_  procedure |
|:-----------------------------------:|--------------------------|--------------------------|--------------------|
| Frequency Tables                    | `freq()`                 | PROC FREQ                | FREQUENCIES        |
| Cross-Tabulations                   | `ctable()`               | PROC FREQ PROC TABULATE  | CROSSTABS          |
| Univariate Statistics (Univariates) | `descr()`                | PROC MEANS PROC SUMMARY  | DESCRIPTIVES       |
| Full-Fledged Dataset   Summaries    | `dfSummary()`            |             ?            |          ?         |


|          Category        | Frequency Tables | Cross-Tabulations        |  Univariate Statistics (Univariates) |
|:------------------------:|------------------|--------------------------|--------------------------------------|
|  _summarytools_ function | `freq()`         | `ctable()`               | `descr()`                            |
|  _SAS_ procedure         | PROC FREQ        |  PROC FREQ PROC TABULATE |  PROC MEANS PROC SUMMARY             |
|  _SPSS_  procedure       | FREQUENCIES      | CROSSTABS                | DESCRIPTIVES                         |


I did find useful a variery of functions spread across several packages, but 
none offered consistency in the presentation, or sometimes even any basic 
aesthetic concern. Moreover, no single package offered all four, nor minimally
all three categories of statistic tables.

The "Full-Fledge Data Summaries" is something I had been working on for some
time. Its first iteration used _Python_'s interface to _SPSS_ objects to 
generate contents similar to what `dfSummary()` offers, minus the graphics. Its
output was directed into _Excel_ spreadsheets which in turn were formatted
using a _VBA_ macro.

The following features from the proprietary software did also influence the
way I approached the development of **summarytools**:

 - **Consistency and flexibility of outputs** : **summarytools** outputs come in 
   two flavours - plain text / markdown, as well as HTML complemented by
   *Twitter Bootstrap Cascading StyleSheets*; individual files can be produced,
   as well as appended.   
 - Split-group analysis : a "quasi-clone" of `by()`, `stby()`, allows creating
   objects which:
     + are based on iterations over groups of data
     + possess attributes specific to **summarytools** inner workings
 - Support for sampling weights 
 - Labelling capabilities : a `label()` function is included in the package,
   although similar functions from the **rapportools** and **Hmisc** packages
   produce the same results when generating **summarytools** tables
 - Control over titles and footnotes : since it is not designed as a standalone
   reporting system, flexibility on that front is limited, but users can rather
   easily:
     + show or hide the heading section that displays data names and a limited
       number of attributes
     + Add captions to plain text / rmarkdown tables, as well as footnotes to
       HTML tables
 - Some more advanced exporting capabilities such as appending results to 
   existing reports
 - Missing vs valid data information



+-------------------+-------------+-------------------------------+---------------------------+
| Type              | Function    | Traditional                   | Tidyverse / pipes         |
+-------------------+-------------+-------------------------------+---------------------------+
| Single Tables     | freq()      | freq(tobacco$gender)          | tobacco %>% freq(gender)  |
|                   |             |                               | tobacco %$% freq(gender)  |
|                   |             |                               | tobacco %>>% freq(gender) |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | descr()     | descr(tobacco$age)            | tobacco %>%               |
|                   |             |                               |   group_by(gender) %>%    |
|                   |             |                               |   freq(smoker)            |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | ctable()    | with(tobacco,                 | tobacco %$%               |
|                   |             |   ctable(gender, smoker)      |   ctable(gender, smoker)  |
|                   |             | )                             |                           |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | dfSummary() | dfSummary(tobacco)            | tobacco %>% dfSummary()   |
|                   |             |                               | tobacco %$% dfSummary()   |
|                   |             |                               | tobacco %>>% dfSummary()  |
+-------------------+-------------+-------------------------------+---------------------------+
| Iteration over    | freq()      | freq(tobacco)                 | tobacco %>% freq()        |
| columns           |             |                               | tobacco %$% freq()        |
|                   |             |                               | tobacco %>>% freq()       |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | ctable()    |                               |                           |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | descr()     | descr(tobacco)                | tobacco %>% descr()       |
|                   |             |                               | tobacco %$% descr()       |
|                   |             |                               | tobacco %>>% descr()      |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | dfSummary() |                               |                           |
+-------------------+-------------+-------------------------------+---------------------------+
| Iteration over    | freq()      | with(tobacco,                 | tobacco %>%               |
| subgroups         |             |   stby(smoker, gender, freq)  |   group_by(gender) %>%    |
|                   |             | )                             |   freq(smoker)            |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | descr()     | with(tobacco,                 | tobacco %>%               |
|                   |             |   stby(age, gender, descr)    |   group_by(gender) %>%    |
|                   |             | )                             |   descr(age)              |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | ctable()    |                               |                           |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | dfSummary() | stby(tobacco,                 | tobacco %>%               |
|                   |             |      tobacco$gender,          |   group_by(gender) %>%    |
|                   |             |      dfSummary)               |   dfSummary()             |
+-------------------+-------------+-------------------------------+---------------------------+
| Two-way iteration | freq()      |                               |                           |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | ctable()    |                               |                           |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | descr()     | stby(tobacco,                 | tobacco %>%               |
|                   |             |      tobacco$gender,          |   group_by(gender) %>%    |
|                   |             |      descr)                   |   descr                   |
+-------------------+-------------+-------------------------------+---------------------------+
|                   | dfSummary() |                               |                           |
+-------------------+-------------+-------------------------------+---------------------------+


# In the Pipeline 

This section demonstrates how to effectively use **summarytools**
with **magrittr** pipe operators and **dplyr**'s `group_by()`.

## Frequency Tables With freq()

Since version 0.9.0, `freq()` works both with vectors/factors and 
data frames. In the latter case, a list-type object is created, just
like using `lapply(df, freq)` would, with a two key distinctions: 1) the 
resulting object's class is _stby_, which allows **summarytools** to 
immediately seize control over their printing or html rendering; 2) a 
threshold is used to decide whether numerical variables should be 
ignored or not, depending on the number of distinct values they contain.
The value of this threshold has a default value of 25 which can be
modified using `st_options(freq.ignore.threshold)`.

### Frequencies for a Single Vector or Factor

<!-- The traditional way: -->

<!-- ```{r, eval=FALSE} -->
<!-- freq(tobacco$smoker) -->
<!-- ``` -->

<!-- The pipes way: -->
<!-- ```{r} -->
<!-- tobacco %>% freq(smoker) -->
<!-- # Also possible: -->
<!-- # tobaaco %$% freq(smoker) -->
<!-- ``` -->

<!-- To obtain grouped statisitics the traditionnal way, we use the `stby()` -->
<!-- function, which works exactly like base R's `by()`: -->
<!-- ```{r, eval=FALSE} -->
<!-- stby(data = tobacco$smoker, INDICES = tobacco$gender, FUN = freq) -->
<!-- ``` -->

<!-- The same can be achieved using `dplyr::group_by`. To avoid a warning about -->
<!-- implicit `NA`'s on the grouping variable, we use `forcats::fct_explicit_na()` -->
<!-- to make them explicit. -->
<!-- ```{r} -->
<!-- library(dplyr) -->
<!-- library(forcats) -->
<!-- tobacco %>%  -->
<!--   group_by(gender = fct_explicit_na(gender)) %>% -->
<!--   freq(smoker, totals = FALSE) -->
<!-- ``` -->

<!-- The `totals = FALSE` argument was added since it is in the spirit of tidy -->
<!-- data. -->

<!-- #### Using parameters to refine results -->
<!-- Here is an example that illustrates the use of parameters; we want the simplest -->
<!-- table showing the 5 most common values for the "disease" variable: -->
<!-- ```{r, eval=FALSE} -->
<!-- freq(tobacco$disease, order = "freq", rows = 1:5,  -->
<!--      report.nas = FALSE, cumul = FALSE) -->
<!-- ``` -->

<!-- ... the same using pipes: -->
<!-- ```{r} -->
<!-- tobacco %>% freq(disease, order = "freq", rows = 1:5,  -->
<!--                  report.nas = FALSE, cumul = FALSE) -->
<!-- ``` -->

<!-- ### Get Really Tidy -->
<!-- The recently introduced function `tb()` takes summarytools objects and turns -->
<!-- them into _real_ tidy tables. The letters "tb" refer to the word / function -->
<!-- _tibble_. It works with `freq()` and `descr()` only, because the structure of -->
<!-- `ctable()` and `dfSummary()` is not compatible with the concept of tidy data. -->

<!-- Here is what a tidied `freq()` table looks like: -->
<!-- ```{r} -->
<!-- tobacco %>% freq(smoker) %>% tbl() -->
<!-- ``` -->

<!-- When one or more grouping variables are added, we have the following results: -->
<!-- ```{r} -->
<!-- tobacco$gender %<>% forcats::fct_explicit_na() -->
<!-- tobacco$age.gr %<>% forcats::fct_explicit_na() -->
<!-- tobacco %>% group_by(gender, age.gr) %>% freq(smoker) %>% tbl() -->
<!-- ``` -->
