---
title: "Custom Statistics in dfSummary"
author: Dominic Comtois
date: "`r Sys.Date()`"
output:
  pdf_document: 
    highlight: tango
    latex_engine: xelatex
    extra_dependencies: xcolor
    includes:
      in_header: 
      - !expr system.file("includes/fig-valign.tex", package = "summarytools")
---

\definecolor{MidnightBlue}{HTML}{2E74B5}

```{r, include=FALSE}
library(knitr)
opts_chunk$set(comment = NA, 
               prompt  = FALSE,
               cache   = FALSE,
               echo    = TRUE,
               results = 'asis')
library(summarytools)
```

This document shows how to customize the content of the *Stats / Values* column
in [\color{MidnightBlue}{data frame
summaries}](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html#data-frame-summaries-dfsummary)
generated using `summarytools::dfSummary()`. This feature was introduced in
version 1.0.0 of
[**summarytools**](https://CRAN.R-project.org/package=summarytools)as a response
to a feature request that came up several times, in a
[\color{MidnightBlue}{form}](https://github.com/dcomtois/summarytools/issues/33)
or
[\color{MidnightBlue}{another}](https://github.com/dcomtois/summarytools/issues/74).

## How it works

Two new options were created: `dfSummary.custom.1` and `dfSummary.custom.2`. The
first one has a predefined value -- it is the one that makes up the fourth row
of the cell (showing IQR and CV). The second one is set to `NA` by default. If
both options are defined (non-`NA`), the cell will now show 5 lines rather than
4, provided there are no additional line feed occurring within the cell, be it
by design or by an "overflow" of one of the custom lines.

### Setup & Baseline

We'll use the first column of *iris* to show results as they are before making
any changes. But first, a little bit of setting-up:

\vspace{8pt}

```{r}
library(summarytools)
st_options(plain.ascii     = FALSE,
           headings        = FALSE,
           footnote        = NA,
           round.digits    = 1,
           style           = "rmarkdown", # For freq(), descr(), & ctable()
           dfSummary.varnumbers   = FALSE,
           dfSummary.valid.col    = FALSE,
           dfSummary.silent       = TRUE,
           dfSummary.style        = "grid",
           tmp.img.dir            = "img")
```

Now let's show the default output: \vspace{8pt}

```{r}
iris_subset <- iris[1]
dfSummary(iris_subset, graph.magnif = .45)
```

## Example 1 : Removing *IQR (CV)*

Setting `dfSummary.custom.1` to `NA` will remove the last line in *Stats /
Values*: \vspace{8pt}

```{r}
st_options(dfSummary.custom.1 = NA)
dfSummary(iris_subset, graph.magnif = .35) # Adjust graph size accordingly
```

## Example 2 : Adding *Q1* & *Q3*

Here we'll create the expression needed to generate new statistics, *Q1* & *Q3*.
The expression is evaluated while looping on column data, and we need to refer
to that data. The variable name to use is, well, `column_data`. Another variable
you can use is `round.digits` (we've set to `1` in the setup chunk on page 1).
\vspace{8pt}

```{r}
st_options(
  dfSummary.custom.1 = 
    expression(
      paste(
        "Q1 - Q3 :",
        round(
          quantile(column_data,
                   probs = .25,
                   type = 2, 
                   names = FALSE,
                   na.rm = TRUE),
          digits = round.digits
        ), " - ",
        round(
          quantile(column_data, 
                   probs = .75,
                   type = 2, 
                   names = FALSE,
                   na.rm = TRUE),
          digits = round.digits
        )
      )
    )           
)

dfSummary(iris_subset, graph.magnif = .45)
```

## Example 3: Inserting Back *IQR (CV)*

It is always possible to reset the value of `dfSummary.custom.1` to its initial
value by using

```{r, eval=FALSE}
st_options(dfSummary.custom.1 = "default")
```

But let's make things a bit more interesting by actually showing *IQR (CV)*
**under** *Q1* & *Q3*. For this, we will use the default expression for
`dfSummary.custom.1` to define `dfSummary.custom.2`:

\vspace{8pt}

```{r}
st_options(
  dfSummary.custom.2 =
    expression(
      paste(
        paste0(
          trs("iqr"), " (", trs("cv"), ") : "
        ),
        format_number(
          IQR(column_data, na.rm = TRUE),
          round.digits
        ),
        " (",
        format_number(
          sd(column_data, na.rm = TRUE) /
              mean(column_data, na.rm = TRUE),
          round.digits
        ),
        ")",
        collapse = "", 
        sep = ""
    )
  )
)

dfSummary(iris[3:5], graph.magnif = .65) # Again, graph size adjusted
```

Don't forget to set `na.rm = TRUE` whenever necessary (most base R statistics
use it with `FALSE` as default).

## Number Formatting

You may have noticed that instead of `round()`, we used `format_number()`, which
is a **summarytools** internal function. It applies not only rounding, but all
relevant formatting attributes as well (*nsmall, decimal.mark, big.mark,
scientific*,and so on).

## Displaying Formatted Expressions

As shown in the
[\color{MidnightBlue}{Introduction to summarytools}](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html)
vignette, the following bit of code can be used to retrieve and format the
expressions stored in the custom options. To achieve good results, the chunk
option `results='markup'` was used for this chunk.

```{r, results='markup'}
st_options(dfSummary.custom.1 = "default")
formatR::tidy_source(
  text   = deparse(st_options("dfSummary.custom.1")),
  indent = 2,
  args.newline = TRUE
)
```

## Useful links

1.  [\color{MidnightBlue}{Introduction to summarytools}](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html)
    (package vignette)
2.  [\color{MidnightBlue}{Summarytools in R Markdown Documents}](https://cran.r-project.org/web/packages/summarytools/vignettes/rmarkdown.html)
    (package vignette)
3.  [\color{MidnightBlue}{Data Frame Summaries in PDF's}](https://raw.githubusercontent.com/dcomtois/summarytools/master/doc/Data-Frame-Summaries-in-PDFs.pdf)
    (supplemental documentation)
