---
title: "Data Frame Summaries in PDF's"
author: Dominic Comtois
date: "`r Sys.Date()`"
output:
  pdf_document:
    highlight: tango
    latex_engine: xelatex
    includes:
      in_header: include-header.tex
papersize: letter
---

\definecolor{MidnightBlue}{HTML}{2E74B5}

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "asis", cache = FALSE)

library(summarytools)
st_options(
  plain.ascii            = FALSE, 
  style                  = "rmarkdown", 
  subtitle.emphasis      = FALSE,
  dfSummary.style        = "grid",
  dfSummary.graph.magnif = .5,
  dfSummary.valid.col    = FALSE,
  dfSummary.silent       = TRUE, 
  tmp.img.dir            = "img"
)
```

Here are the instructions for setting up [\color{MidnightBlue}{\emph{R
Markdown}}](https://rmarkdown.rstudio.com/) documents in order to generate *pdf*
documents with [\color{MidnightBlue}{data frame
summaries}](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html#data-frame-summaries-dfsummary)
(`summarytools::dfSummary()`) that contain images.

# 1. The Graphics Alignment Problem {#problem}

Although generating *html* or *Word* documents from *Rmd*'s containing
`dfSummary()` outputs is a smooth and painless process, there is a major problem
when it comes to generating *pdf*'s. The graphs, instead of being vertically
centered, appear as though they were sitting on top of all the other cells'
content:

```{r}
dfSummary(iris[3:5], headings = FALSE)
```

\input{include-renew-cmd.tex}

# 2. The Solution

To correct this issue, we need to redefine the `\includegraphics` command. This
can be done in multiple ways, but the simplest is to include in your document's
header a *tex* file which is designed to do just that. It can be achieved by
configuring the YAML section as follows.

## 2.1 The YAML Section

        ---
        title: "My Own Private PDF"
        output:
          pdf_document:
            highlight: tango
            latex_engine: xelatex
            includes:
              in_header: 
              - !expr system.file("includes/fig-valign.tex",
                                  package = "summarytools")
        papersize: letter
        ---

The solution presented here requires that some *tex* code be included in the
YAML section of the Rmd document. You can use your own *tex* file, or use the
one that is part of the package as of version 1.0.0 (July 2021). and include it
in from the YAML section using `system.file()`.

The `latex_engine: xelatex` part is not mandatory for the solution to work. But
there are several advantages to using it; I use it systematically and see only
advantages to it, so I can only advise you do the same.

This solution is not perfect; if your *pdf* document relies on the use of
`\includegraphics` in other sections, you might notice newly *mis*aligned
images. Thankfully, there is a way to go around this (see
[\color{MidnightBlue}{section 2.3}](#robust)).

### Using Your Own *tex* File

If you prefer including your own *tex* file, here is what it should (minimally)
contain:

        \usepackage{graphicx}
        \usepackage[export]{adjustbox}
        \usepackage{letltxmacro}
        \LetLtxMacro{\OldIncludegraphics}{\includegraphics}
        \renewcommand{\includegraphics}[2][]{\raisebox{0.5\height}%
          {\OldIncludegraphics[valign=t,#1]{#2}}}

The only impact on your YAML section will be the `in_header:` attribute which
will need to point to this file, using an absolute or relative path. If the file
is kept in the same directory as your *Rmd* document, you'll use
`in_header: fig-align.tex` (supposing you use that file name).

## 2.2 Example

Here is a setup chunk, followed by a call to `dfSummary()`.

```{r, eval=FALSE}
library(summarytools)
st_options(
  plain.ascii            = FALSE, 
  subtitle.emphasis      = FALSE,
  style                  = "rmarkdown", # For other summarytools objects (freq, descr...)
  dfSummary.style        = "grid",
  dfSummary.graph.magnif = .5,
  dfSummary.valid.col    = FALSE,
  tmp.img.dir            = "/tmp"  # Recommended on Linux/OS X; On 
                                   # Windows, "img" is suggested 
)
```

 

Now that the setup is done, we can generate the results.

 

```{r}
define_keywords(title.dfSummary = "Data Frame Summary in PDF Document")
dfSummary(iris[3:5])
```

# 3. A More Robust Solution

If redefining the `\includegraphics` command causes problems elsewhere in your
document[^1], following these instructions should take care of it (file names
and location are suggestions only):

[^1]: There must be a *law of conservation of brokenness* sitting somewhere
    waiting to be discovered (although one could argue that it is merely a
    corollary to
    [\color{MidnightBlue}{Murphy's law}](https://en.wikipedia.org/wiki/Murphy%27s_law))

1.  Split the contents of `fig-valign.tex` into two files in your *Rmd*
    document's directory:

    i.  `load-pkgs.tex` -- contains only the first three lines (the
        `\usepackage` commands only)

    ii. `renew-cmd.tex` -- contains the remaining lines, which store the
        existing `\includegraphics` command as a macro and redefine it

2.  Include the first file in the YAML section (`in_header: load-pkgs.tex`)

3.  Before the `dfSummary()` chunk(s), paste this *tex* command on a new line:

        \input{renew-cmd.tex}

4.  After the chunk(s), set the `\includegraphics` back to its original value
    using the following command (also on a new line):

        \let\includegraphics\OldIncludegraphics

You might need to repeat steps 3 and 4 several times if your document alternates
between `dfSummary()` tables and other content with images.

### Proof That `includegraphics` Is Restored to Original

\let\includegraphics\OldIncludegraphics

At this stage, the `\let\includegraphics\OldIncludegraphics` *tex* command has
been executed.

 

```{r}
dfSummary(iris[5], headings = FALSE)
```

If the operation of restoring the command worked, the results should be back to
being misaligned, just as they were in the
[\color{MidnightBlue}{very first section}](#problem).

## Closing Remarks

If you are a \LaTeX  guru and can think of a simpler solution, please do let me
know either by opening an
[\color{MidnightBlue}{issue}](https://github.com/dcomtois/summarytools/issues)
or by sending me an email[^2]. my address is available in the
[\color{MidnightBlue}{package's GitHub
page}](https://github.com/dcomtois/summarytools) as well as in the
[\color{MidnightBlue}{package's auto-generated pdf manual}](https://cran.r-project.org/web/packages/summarytools/summarytools.pdf).

[^2]: My email address is available in the
    [\color{MidnightBlue}{package's GitHub
    page}](https://github.com/dcomtois/summarytools) as well as in the
    [\color{MidnightBlue}{package's auto-generated pdf manual}](https://cran.r-project.org/web/packages/summarytools/summarytools.pdf).

## Useful links:

1.  [\color{MidnightBlue}{Introduction to summarytools}](https://cran.r-project.org/web/packages/summarytools/vignettes/introduction.html)
    (package vignette)
2.  [\color{MidnightBlue}{Summarytools in R Markdown Documents}](https://cran.r-project.org/web/packages/summarytools/vignettes/rmarkdown.html)
    (package vignette)
3.  [\color{MidnightBlue}{Custom Statistics in dfSummary}](https://raw.githubusercontent.com/dcomtois/summarytools/master/doc/Custom-Statistics-in-dfSummary.pdf)
    (supplemental documentation)
4.  [\color{MidnightBlue}{This StackOverflow question}](https://stackoverflow.com/questions/5845887/how-do-i-use-renewcommand-to-get-back-my-greek-letters)
    provides an additional example of how to revert a renewed command back to
    its original value.
